---
title: "Estimating years of life lost in Scotland and the rest of the UK from Covid-19 under different scenarios"
output: html_notebook
---

# Software prereqs 

```{r}
pacman::p_load(here, tidyverse, readxl)

```


# Background

The aim of this document is to provide some reasonable estimates of the likely effect that a covid-19 outbreak will have on mortality in 2020 conditional on a range of assumptions. 

This will involve comparing various 'covid' scenarios against a 'control' scenario, in terms of the numbers of deaths expected in the country at different ages. Some important key parameters: 

* The ratio of true cases to confirmed cases 
* The proportion of the population who will be infected in 2020. 
* The hazard multiplier on age-sex specific mortality rates conditional on being a confirmed case. 

Many of these are not known yet and so a range of different scenarios should be considered. 

To start with we want the following data: 

1. Age-sex population structure of Scotland and other UK nations in 2020 or recent year 
2. Age-sex specific mortality rates (12 month hazards)

Let's start with just Scotland. Soon we'll want to do for other countries and to create an app that would allow different populations to be selected. 

# Scotland population structure 

Let's use the 2018 lifetables from the ONS I used previously as part of the [Bayes Factor Slowdown](https://github.com/JonMinton/bayes_factor_slowdown/) repo.

```{r}
lifetables <- read_rds(here("data", "lifetables", "all_ons_lifetables_tidied.rds"))

```


For the last year, and for Scotland 

```{r}
lt_scotland <- 
  lifetables %>% 
    group_by(population) %>% 
    filter(year == max(year)) %>% 
    filter(population == "Scotland")
```

Now population structure for last year.

I've downloaded the data from [here](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/analysisofpopulationestimatestool) and put it in the data directory.  

The data required are in the MYEB2 tab

```{r}
pop_uk <- 
  read_excel(here("data", "theanalysisofpopulationestimatestool2018.xlsx"),
             sheet = "MYEB2"
             )
```


```{r}
pop_uk %>% 
  count(code)

```

This doesn't have Scotland! 

For Scotland I'm going to use [this page](https://www.nrscotland.gov.uk/statistics-and-data/statistics/statistics-by-theme/population/population-estimates/mid-year-population-estimates/mid-2018) and [this file](https://www.nrscotland.gov.uk/files//statistics/population-estimates/mid-18/mid-year-pop-est-18-tabs.zip). The file is a zip file so I'm going to download it and save as before. 


Update: now going to use [the resource I used](https://www.opendata.nhs.scot/dataset/7f010430-6ce1-4813-b25c-f7f335bdc4dc/resource/c505f490-c201-44bd-abd1-1bd7a64285ee/download/dz2011-pop-est_30082019.csv) for the geovariation project 


```{r}

pop_scot_loc <- "https://www.opendata.nhs.scot/dataset/7f010430-6ce1-4813-b25c-f7f335bdc4dc/resource/c505f490-c201-44bd-abd1-1bd7a64285ee/download/dz2011-pop-est_30082019.csv"
(dta_scot <- read_csv(pop_scot_loc))

```

```{r}
(
dta_scot_2018 <- 
  dta_scot %>% 
    set_names(nm = tolower(names(.))) %>% 
    mutate(sex = tolower(sex)) %>% 
    filter(dz2011 == "S92000003") %>% 
    filter(year == max(year)) %>% 
    select(-allages, -dz2011, -dz2011qf, -year) %>% 
    gather(key = "age", value = N,  -sex) %>% 
    mutate(age = str_remove(age, "^age") %>% str_remove("plus$") %>% as.numeric()) 
)
```

So this gives the number of people at each age and sex in Scotland. 

With these two pieces of information we can start to produce estimates of the numbers of expected deaths under 2018 mortality hazards, and for a range of covid scenarios. 

Baseline expected deaths 

```{r}

lt_scotland %>% 
  select(x, sex, mx) %>% 
  inner_join(dta_scot_2018, by = c("x" = "age", "sex" = "sex")) %>%
  ungroup() %>% 
  select(-population) %>% 
  mutate(n_control = mx * N)
```

Northern Ireland

[This](https://www.nisra.gov.uk/publications/2018-mid-year-population-estimates-northern-ireland) website appears to have it. I've downloaded the excel file to data/population (MYE18_SYA.xlsx).

```{r}
(
  ni_pop <- read_excel(here("data", "population", "MYE18_SYA.xlsx"), sheet = "Flat")
)
```

```{r}

ni_pop %>% 
  filter(area == "1. Northern Ireland") %>% 
  filter(year == max(year))  %>% 
#  count(gender) %>% 
  mutate(
    sex = case_when(
      gender == "Males"        ~ "male",
      gender == "Females"      ~ "female",
      TRUE                     ~ NA_character_
    )
  ) %>% 
  filter(!is.na(sex)) %>% 
  select(sex, age, N = MYE)

```


Now to produce a function for various scenarios

```{r}
adjust_Mx <- function(df, truePrevalence, ratioTrueToConfirmed, hrConfirmed, hrUnconfirmed = 1){
  
  share_uninfected  <- 1 - truePrevalence
  share_confirmed   <- truePrevalence / ratioTrueToConfirmed
  share_unconfirmed <- 1 - share_uninfected - share_confirmed
  
  df %>% 
    mutate(mx_adjusted = 
      (mx * share_uninfected)                    + 
      ((mx * hrUnconfirmed) * share_unconfirmed) +
      ((mx * hrConfirmed)   * share_confirmed)
  ) %>% 
  mutate(
    n_control = mx * N,
    n_covid = mx_adjusted * N
  )
}



```


Let's try this with an HR conditional on confirmed covid of 2.0, a 60% prevalence, and a 10:1 ratio of unconfirmed to true cases 

```{r}
(df_firstscenario <-
  lt_scotland %>% 
    select(x, sex, mx) %>% 
    inner_join(dta_scot_2018, by = c("x" = "age", "sex" = "sex")) %>%
    ungroup() %>% 
    select(-population) %>% 
    adjust_Mx(truePrevalence = 0.6, hrConfirmed = 2.0, ratioTrueToConfirmed = 10)
)
```

Now a suitable visualisation 

```{r}
df_cumulative <- 
  df_firstscenario %>% 
    group_by(sex) %>% 
    arrange(x) %>% 
    mutate(
      cumulative_n_control = cumsum(n_control),
      cumulative_n_covid   = cumsum(n_covid)
    ) %>% 
    ungroup() %>% 
    mutate(
      cumulative_diff = cumulative_n_covid - cumulative_n_control
    ) %>% 
    select(x, sex, cumulative_n_control, cumulative_n_covid, cumulative_diff) %>% 
    gather(-x, -sex, key = "key", value = "n") %>% 
    mutate(key = factor(key, levels = c("cumulative_n_control", "cumulative_n_covid", "cumulative_diff")))

df_cumulative %>% 
  ggplot(aes(x = x, y = n, group = key , colour = key, shape = key)) + 
  facet_wrap(~ sex) + 
  geom_line() +
  labs(x = "Age in single years", y = "Number of deaths")

df_cumulative %>% 
  group_by(sex) %>% 
  filter(x == max(x)) %>% 
  select(-x) %>% 
  spread(key, n)

```

Let's now build a simple app for looking at different scenarios

Populations:

* Scotland
* Wales
* England 


